Evaluation of 3-point correlation functions from structural images on CPU and GPU architectures: accounting for anisotropy effects

### 2024-09-03 paper.tex
Fix isosceles
Fix another image
Fix formula
Replace \times

The second approach is to compute the whole correlation map in the Fourier
domain. Suppose we have a function
$f: \mathbb{R} \rightarrow \mathbb{R}$. Triple correlation for this function is:
\begin{equation}
  g(t_1, t_2) = \int f(\tau) f(\tau + t_1) f(\tau + t_2) d \tau.
\end{equation}
We can use a well-known formula to compute Fourier image of $g$:
\begin{equation}
  \hat{g}(z_1, z_2) = \hat{f}(z_1) \hat{f}(z_2) \overline{\hat{f}(z_1 + z_2)}
\end{equation}
This algorithm has a better computational complexity compared with the previous
algorithms if the whole correlation map is needed ($O(n^2 \log n)$ vs
$O(n^3)$ where $n$ is a number of elements in the input array). The whole
correlation map, however, requires a lot of memory and is rarely
needed. Imagine, that the input array $A$ is of dimensions $500 \times 500$.
Then, considering that single precision floating point numbers are used,
$4 \cdot 2 \cdot 500^4 / 1000^3 = 500$ GB are needed to store Fourier image of
the map. Moreover, Fourier image of the whole map has to be computed even if only one
element of the map in spatial domain is required. The aforementioned limitations
render this approach impractical.

@@

To compute $C_3$ function we need to use a connected-component labeling
algorithm \cite{4728561,PhysRevB.14.3438} to label clusters. Then, again, a
slightly modified algorithm for $S_3$ is used to assess cluster function:
\begin{algorithmic}[1]
  \Procedure{$C_3$}{$A, n, x_1, x_2$}
  \State $A_n \gets I_n (A)$
  \State $C_n \gets \mathfrak{C}(A_n)$
  \State $C'_n \gets \mathfrak{S}_{x_1}(C_n)$
  \State $C''_n \gets \mathfrak{S}_{x_2}(C_n)$
  \State $Tmp \gets C_n \odot C'_n \odot C''_n$
  \State \textbf{return} $Sum(Tmp) / Norm(A, x_1, x_2)$
  \EndProcedure
\end{algorithmic}
Here $\mathfrak{C}$ is connected-component labeling operator and $\odot$ is
defined as follows:
\begin{equation}
  x \odot y = \left\{
  \begin{array}{ll}
    1 & \quad x = y \ne L_{(void)} \\
    0 & \quad \text{otherwise}.
  \end{array}
  \right.
\end{equation}
where $L_{(void)}$ is a label of the void phase.

@@

To detect the direction of anisotropy we use an $N \times N$ (where $N$ is
dimensionality of the sample) covariance matrix
$\Sigma = \left\{ \sigma_{ij} \right\}$ for coordinates of points in the phase of
interest:
\begin{equation}
  \sigma_{ij} = \sum_{x \in \left\{
    \begin{array}{l}
      \text{Coordinates of points} \\
      \text{in the phase of interset}
    \end{array}
    \right\}} (x_i - \overline{x_i})(x_j - \overline{x_j})
\end{equation}
and decompose $\Sigma$ as $T \Lambda T^{-1}$ where $T$ is a unitary matrix and
$\Lambda$ is diagonal. $T$ can always be made so that it defines a rotation (by
multiplication of one of the columns by $-1$ if needed, so that
$\det T = 1$). Application of this rotation to the sample (or its inverse to the
pattern) makes the direction of anisotropy coaligned with the pattern, so
anisotropy can be better represented in the correlation function. For the sake
of visual clarity in this study we choose to rotate the image instead of the
pattern (\cref{fig:aniso}). Any rotation can potentially reduce accuracy of
computed functions due to necessity to interpolate the image during
rotation, and this important issue is explored later.
\begin{figure*}[ht]
  \centering
  \includegraphics[width=0.8\linewidth]{images/fig_rot.png}
  \caption{Examples of directional analysis of images with the help of $S_3$
    function when dealing with highly anisotropic media. The right-triangular
    pattern (\cref{fig:pattern}, right side) which we use in our algorithm is
    most sensitive to anisotropy which is directed along one of the
    axes. \\ From top down, left to right: an original image of an anisotropic
    pore; a rotated image of the pore, so it is positioned along the major axis;
    $S_3$ function for the original pore; $S_3$ function for the rotated pore
    with a vector $(0.92, -0.4)$ being a direction of elongation of the pore;
    $S_3$ function averaged across rotations similar to previously known
    algorithms \cite{berryman1988,SMITH1988176,malmir2018}.}
  \label{fig:aniso}
\end{figure*}

@@

The results of the comparison are shown in \cref{fig:comparison_patterns}. For
isotropic samples curves for directional and ensemble averaged, as well as for
right-angle isosceles triangle, basically collapse to each other with minimal
differences.
\begin{figure*}[ht]
  \centering \includegraphics[width=0.8\linewidth]{images/fig_compar.png}
  \caption[]{A comparison of computations of $S_3$ correlation function for the
    Sandstone sample using different patterns with and without random
    rotations. Left: evaluated with isosceles triangle pattern with the side $r$
    and the angle between edges $\phi$; sampling triangles are aligned with the
    plane XY.  Middle: evaluated with equilateral triangle pattern with the side
    $r$; sampling triangles are aligned with the plane XY.  Right: evaluated
    with equilateral triangle pattern with the side $r$. The result is averaged
    over 20 random rotations of the sampling triangle. Lower image: comparison
    of $S_3$ functions calculated with right triangle pattern with length of the
    both catheti $r$ and equilateral triangle pattern (both averaged over
    rotations and aligned with the plane XY).}
  \label{fig:comparison_patterns}
\end{figure*}

@@

\subsection{Applications to material images}
\begin{table*}[!htp]
  \centering
  \begin{tabular}{|c|c|c|c|c|}
    \hline
    Sample name & Image type & Dimensions (pixels/voxels) & Image resolution ($\mu m$)
    & $Q_{0.5}$ \\
    \hline
    Sandstone & XCT &  $500 \times 500 \times 500$ & 2.25 & 0.9703 \\
    Soil & Reconstruction & $500 \times 500 \times 500$ & 0.72 & 0.9431 \\
    Shale & Reconstruction & $160 \times 160 \times 160$ & 2.25 & 0.8912 \\
    Sandstone & SEM &  $1280 \times 869$ & 0.10 & 0.9613 \\
    \hline
  \end{tabular}
  \caption{Summary of samples used for calculation of correlation functions.}
  \label{tab:summary}
\end{table*}
In this section we compute correlation functions for 2- and 3-dimensional porous
media samples. Table \ref{tab:summary} contains information about these
samples. Note that we compute surface correlation functions only for those
samples for which the criterion $Q_{0.5} > 0.97$ holds (the two sandstones and
the soil) \cite{samarin2023robust}. Images of the samples are presented on
\cref{fig:samples}. On plots of correlation functions for isotropic samples we
provide a family of curves with the first argument of the correlation function
being fixed. For anisotropic samples we provide heatmaps of correlation
functions where anisotropy can be easily observed.
\begin{figure*}[tp]
  \centering
  \subfigure[Sanstone XCT]{
    \includegraphics[width=0.4\linewidth]{images/15581.png}
    \label{fig:sample-sandstone-xct}}
  \hfill
  \subfigure[Sandstone SEM (different clusters of void phase are shown in color)]{
    \includegraphics[width=0.4\linewidth, frame]{images/sandstone1.png}
    \label{fig:sample-sandstone}}
  \vskip\baselineskip
  \subfigure[Soil]{
    \includegraphics[width=0.4\linewidth]{images/soil.png}
    \label{fig:sample-soil}}
  \hfill
  \subfigure[Shale]{
    \includegraphics[width=0.4\linewidth]{images/shale.png}
    \label{fig:sample-shale}}
  \caption[]{Images of samples used for calculation of higher-order correlation
    functions.}
  \label{fig:samples}
\end{figure*}

@@

To estimate the error introduced by rotation, consider a two-phase medium
\cref{fig:roterror-noise} whose void phase is defined by an inequality
$f(x, y, z) > U$ where $f$ is a value noise function
\cite[Section IV.B]{samarin2023robust} and $U$ is a threshold. We produce images
of the noise rotated around $Z$ axis by $\theta$ radians by discretizing an
inequation $f(p(x'), p(y'), p(z')) > U$ where:
\begin{equation*}
  \begin{aligned}
    x' &= x \cos \theta + y \sin \theta \\
    y' &= y \cos \theta - x \sin \theta \\
    z' &= z
  \end{aligned}
\end{equation*}
and
\begin{equation*}
  \begin{aligned}
    p(x) &= \left\{
    \begin{array}{ll}
      x'     & \quad x' < 1 \\
      2 - x' & \quad \text{otherwise}
    \end{array}
    \right. \\
    x ' &= x \mod 2.
  \end{aligned}
\end{equation*}
Therefore we assume the noise to be periodic with the period equal to $2$ and
the interface between phases is continuous. Discretizing this noise for
different values of $\theta \in [0, \pi/4]$ with two different resolutions we
obtain the first dataset which contains the best possible discretization of the
rotated noise at those resolutions. For this study we have produced images with
dimensionality $300 \times 300 \times 300$ and $600 \times 600 \times 600$
voxels (i.e., the second resolution is twice as much as the first).

@@

Plots of maximal relative errors for functions $S_3$ and $F_{sss}$ are on
\cref{fig:roterror-error}. When calculating the error of the
surface-surface-surface function we omit values $F_{sss}(s_1, s_2)$ if either of
its arguments is less than 10 voxels because this function is undefined for
either $s_1$ or $s_2$ being equal to zero and our edge detection filter gives at
least 7 voxel transition zone where the correlation function is affected by this
indeterminacy. The maximal error for all $\theta$'s are 0.6\% for $S_3$ and 5.0\%
for $F_{sss}$ for dimensionality $300 \times 300 \times 300$ and 0.3\% for $S_3$
and 4.5\% for $F_{sss}$ for dimensionality $600 \times 600 \times 600$.

>>>>

The second approach is to compute the whole correlation map in the Fourier
domain. Suppose we have a function
$f: \mathbb{R} \rightarrow \mathbb{R}$. Triple correlation for this function is:
\begin{equation}
  g(t_1, t_2) = \int f(\tau) f(\tau + t_1) f(\tau + t_2) d \tau.
\end{equation}
We can use a well-known formula to compute Fourier image of $g$:
\begin{equation}
  \hat{g}(z_1, z_2) = \hat{f}(z_1) \hat{f}(z_2) \overline{\hat{f}(z_1 + z_2)}
\end{equation}
This algorithm has a better computational complexity compared with the previous
algorithms if the whole correlation map is needed ($O(n^2 \log n)$ vs
$O(n^3)$ where $n$ is a number of elements in the input array). The whole
correlation map, however, requires a lot of memory and is rarely
needed. Imagine, that the input array $A$ is of dimensions $500^2$. Then,
considering that single precision floating point numbers are used,
$4 \cdot 2 \cdot 500^4 / 1000^3 = 500$ GB are needed to store Fourier image of
the map. Moreover, Fourier image of the whole map has to be computed even if
only one element of the map in spatial domain is required. The aforementioned
limitations render this approach impractical.

@@

To compute $C_3$ function we need to use a connected-component labeling
algorithm \cite{4728561,PhysRevB.14.3438} to label clusters. Then, again, a
slightly modified algorithm for $S_3$ is used to assess cluster function:
\begin{algorithmic}[1]
  \Procedure{$C_3$}{$A, n, x_1, x_2$}
  \State $A_n \gets I_n (A)$
  \State $C_n \gets \mathfrak{C}(A_n)$
  \State $C'_n \gets \mathfrak{S}_{x_1}(C_n)$
  \State $C''_n \gets \mathfrak{S}_{x_2}(C_n)$
  \State $Tmp \gets C_n \odot C'_n \odot C''_n$
  \State \textbf{return} $Sum(Tmp) / Norm(A, x_1, x_2)$
  \EndProcedure
\end{algorithmic}
Here $\mathfrak{C}$ is connected-component labeling operator and $\odot$ is
defined as follows:
\begin{equation}
  x \odot y = \left\{
  \begin{array}{ll}
    1 & \quad x = y \ne B \\
    0 & \quad \text{otherwise}.
  \end{array}
  \right.
\end{equation}
where $B$ is a background label (i.e. a label for all elements which are not in
the phase $n$).

@@

To detect the direction of anisotropy we use an $N \times N$ (where $N$ is
dimensionality of the sample) covariance matrix
$\Sigma = \left\{ \sigma_{ij} \right\}$ for coordinates of points in the phase of
interest:
\begin{equation}
  \sigma_{ij} = \sum_{x \in \left\{
    \begin{array}{l}
      \text{Coordinates of points} \\
      \text{in the phase of interset}
    \end{array}
    \right\}} (x_i - \overline{x_i})(x_j - \overline{x_j})
\end{equation}
and decompose $\Sigma$ as $T \Lambda T^{-1}$ where $T$ is a unitary matrix and
$\Lambda$ is diagonal. $T$ can always be made so that it defines a rotation (by
multiplication of one of the columns by $-1$ if needed, so that
$\det T = 1$). Application of this rotation to the sample (or its inverse to the
pattern) makes the direction of anisotropy coaligned with the pattern, so
anisotropy can be better represented in the correlation function. For the sake
of visual clarity in this study we choose to rotate the image instead of the
pattern (\cref{fig:aniso}). Any rotation can potentially reduce accuracy of
computed functions due to necessity to interpolate the image during
rotation, and this important issue is explored later.
\begin{figure*}[ht]
  \centering
  \subfigure[Original image]{
    \includegraphics[width=0.4\linewidth,frame]{images/aniso.png}}
  \hfill
  \subfigure[Rotated image]{
    \includegraphics[width=0.4\linewidth,frame]{images/aniso-rot.png}}
  \vskip\baselineskip
  \includegraphics[width=0.3\linewidth]{images/aniso-s3.png}
  \hfill
  \includegraphics[width=0.3\linewidth]{images/aniso-rot-s3.png}
  \hfill
  \includegraphics[width=0.3\linewidth]{images/aniso-avg-s3.png}
  \caption{Examples of directional analysis of images with the help of $S_3$
    function when dealing with highly anisotropic media. The right-triangular
    pattern (\cref{fig:pattern}, right side) which we use in our algorithm is
    most sensitive to anisotropy which is directed along one of the
    axes. \\ From top down, left to right: an original image of an anisotropic
    pore; a rotated image of the pore, so it is positioned along the major axis;
    $S_3$ function for the original pore; $S_3$ function for the rotated pore
    with a vector $(0.92, -0.4)$ being a direction of elongation of the pore;
    $S_3$ function averaged across rotations similar to previously known
    algorithms \cite{berryman1988,SMITH1988176,malmir2018}.}
  \label{fig:aniso}
\end{figure*}

@@

The results of the comparison are shown in \cref{fig:comparison_patterns}. For
isotropic samples curves for directional and ensemble averaged, as well as for
right-angle isosceles triangle, basically collapse to each other with minimal
differences.
\begin{figure*}[ht]
  \centering
  \subfigure[Isosceles with right angle]{
    \includegraphics[width=0.4\linewidth]{images/isosceles.png}}
  \hfill
  \subfigure[Equilateral]{
    \includegraphics[width=0.4\linewidth]{images/equilateral.png}}
  \vskip\baselineskip
  \subfigure[With random rotations]{
    \includegraphics[width=0.4\linewidth]{images/equilateral-avg.png}}
  \hfill
  \subfigure[Comparison plot]{
    \includegraphics[width=0.4\linewidth]{images/alltogether.png}}
  \caption[]{A comparison of computations of $S_3$ correlation function for the
    Sandstone sample using different patterns with and without random
    rotations. Left: evaluated with isosceles triangle pattern with the side $r$
    and the angle between edges $\phi$; sampling triangles are aligned with the
    plane XY.  Middle: evaluated with equilateral triangle pattern with the side
    $r$; sampling triangles are aligned with the plane XY.  Right: evaluated
    with equilateral triangle pattern with the side $r$. The result is averaged
    over 20 random rotations of the sampling triangle. Lower image: comparison
    of $S_3$ functions calculated with right triangle pattern with length of the
    both catheti $r$ and equilateral triangle pattern (both averaged over
    rotations and aligned with the plane XY).}
  \label{fig:comparison_patterns}
\end{figure*}

@@

\subsection{Applications to material images}
\begin{table*}[!htp]
  \centering
  \begin{tabular}{|c|c|c|c|c|}
    \hline
    Sample name & Image type & Dimensions (pixels/voxels) & Image resolution ($\mu m$)
    & $Q_{0.5}$ \\
    \hline
    Sandstone & XCT &  $500^3$ & 2.25 & 0.9703 \\
    Soil & Reconstruction & $500^3$ & 0.72 & 0.9431 \\
    Shale & Reconstruction & $160^3$ & 2.25 & 0.8912 \\
    Sandstone & SEM &  $1280 \times 869$ & 0.10 & 0.9613 \\
    \hline
  \end{tabular}
  \caption{Summary of samples used for calculation of correlation functions.}
  \label{tab:summary}
\end{table*}
In this section we compute correlation functions for 2- and 3-dimensional porous
media samples. Table \ref{tab:summary} contains information about these
samples. Note that we compute surface correlation functions only for those
samples for which the criterion $Q_{0.5} > 0.97$ holds (the two sandstones and
the soil) \cite{samarin2023robust}. Images of the samples are presented on
\cref{fig:samples}. On plots of correlation functions for isotropic samples we
provide a family of curves with the first argument of the correlation function
being fixed. For anisotropic samples we provide heatmaps of correlation
functions where anisotropy can be easily observed.
\begin{figure*}[tp]
  \centering
  \subfigure[Sanstone XCT]{
    \includegraphics[width=0.4\linewidth]{images/15581.png}
    \label{fig:sample-sandstone-xct}}
  \hfill
  \subfigure[Sandstone SEM (different clusters of void phase are shown in color)]{
    \includegraphics[width=0.4\linewidth, frame]{images/sandstone1.png}
    \label{fig:sample-sandstone}}
  \vskip\baselineskip
  \subfigure[Soil]{
    \includegraphics[width=0.4\linewidth]{images/soil.png}
    \label{fig:sample-soil}}
  \hfill
  \subfigure[Shale]{
    \includegraphics[width=0.4\linewidth]{images/shale.png}
    \label{fig:sample-shale}}
  \caption[]{Images of samples used for calculation of higher-order correlation
    functions.}
  \label{fig:samples}
\end{figure*}

@@

To estimate the error introduced by rotation, consider a two-phase medium
\cref{fig:roterror-noise} whose void phase is defined by an inequality
$f(x, y, z) > U$ where $f$ is a value noise function
\cite[Section IV.B]{samarin2023robust} and $U$ is a threshold. We produce images
of the noise rotated around $Z$ axis by $\theta$ radians by discretizing an
inequation $f(p(x'), p(y'), p(z')) > U$ where:
\begin{equation*}
  \begin{aligned}
    x' &= x \cos \theta + y \sin \theta \\
    y' &= y \cos \theta - x \sin \theta \\
    z' &= z
  \end{aligned}
\end{equation*}
and
\begin{equation*}
  \begin{aligned}
    p(x) &= \left\{
    \begin{array}{ll}
      x'     & \quad x' < 1 \\
      2 - x' & \quad \text{otherwise}
    \end{array}
    \right. \\
    x ' &= x \mod 2.
  \end{aligned}
\end{equation*}
Therefore we assume the noise to be periodic with the period equal to $2$ and
the interface between phases is continuous. Discretizing this noise for
different values of $\theta \in [0, \pi/4]$ with two different resolutions we
obtain the first dataset which contains the best possible discretization of the
rotated noise at those resolutions. For this study we have produced images with
dimensionality $300^3$ and $600^3$ voxels (i.e., the second resolution is twice
as much as the first).

@@

Plots of maximal relative errors for functions $S_3$ and $F_{sss}$ are on
\cref{fig:roterror-error}. When calculating the error of the
surface-surface-surface function we omit values $F_{sss}(s_1, s_2)$ if either of
its arguments is less than 10 voxels because this function is undefined for
either $s_1$ or $s_2$ being equal to zero and our edge detection filter gives at
least 7 voxel transition zone where the correlation function is affected by this
indeterminacy. The maximal error for all $\theta$'s are 0.6\% for $S_3$ and 5.0\%
for $F_{sss}$ for dimensionality $300^3$ and 0.3\% for $S_3$ and 4.5\% for
$F_{sss}$ for dimensionality $600^3$.
